Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-06-03 paper/main.tex
new numbers (exept for patterns-delag)

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parametrized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 366 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G}) from the Spot library~\cite{duret.16.atva2}
and \texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA
by \texttt{ltl3tela} with option \texttt{-D1} and by another two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}. Some translators failed on some
formulas: it usually reached a timeout of 60 seconds or it produced an
automaton that cannot be parsed by the Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions of the
translators. For each set of formulas and for each translator, it also
provides the number of fails, the number of produced automata with
less than two marks, and the number of automata with at least two
marks. The numbers of automata with at least two marks are typeset in
bold as these automata are actually used for the experimental
evaluation of our reduction technique.

@@

\begin{table}[t]
  \caption{The cumulative numbers of acceptance marks before and after
    reduction for various reduction levels and automata sets given by
    the translator and formula set. The column \emph{reduction} shows
    the cumulative percentage of saved acceptance marks and
    \emph{time} reports the cumulative reduction time. The column
    \textit{solver timeout} indicates the number of instances where
    the last call to the QBF solver timed out.  }
\label{tab:cumm-data}
\centering
\setlength{\tabcolsep}{3.9pt}
\newcommand{\mybox}[1]{\fbox{\begin{minipage}{4.7cm}\centering#1\end{minipage}}}
\begin{tabular}{lcrrrrrrrr}
  \toprule
  && \multicolumn{4}{c}{\makecell{reduction of marks in automata\\[-.8ex] from formulas from literature}}
  & \multicolumn{4}{c}{\makecell{reduction of marks in automata\\[-.8ex] from random formulas}}
  \\ \cmidrule(lr){3-6} \cmidrule(lr){7-10}
  translator & \makecell{reduction\\[-1ex] level}
  & marks & \makecell{reduction\\[-1ex] [\%]} & \makecell{time\\[-1ex] [s]} & \makecell{solver\\[-1ex] timeout} 
  & marks & \makecell{reduction\\[-1ex] [\%]} & \makecell{time\\[-1ex] [s]} & \makecell{solver\\[-1ex] timeout} \\
  \midrule
%  \multirow{5}{*}{\makecell{\texttt{ltl2tgba -G}}} &
  \multicolumn{2}{l}{\texttt{ltl2tgba -G}} 
  & \multicolumn{4}{c}{\mybox{198 marks in 70 automata}}
  & \multicolumn{4}{c}{\mybox{XYZ marks in XYZ automata}}\\
  & 1
  & 198 & 0.0~~~ & 49.3 & 0~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                     
  & 2
  & 198 & 0.0~~~ & 78.9 & 0~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                     
  & 3
  & 189 & 4.5~~~ & 1222.8 & 7~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                     
  & multi
  & 189 & 4.5~~~ & 1259.1 & 6~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                    
  \midrule
%  \multirow{5}{*}{\makecell{\texttt{ltl2tgba -G}}} &
  \multicolumn{2}{l}{\texttt{ltl3tela}} 
  & \multicolumn{4}{c}{\mybox{348 marks in 91 automata}}
  & \multicolumn{4}{c}{\mybox{XYZ marks in XYZ automata}}\\
  & 1
  & 332 & 4.6~~~ & 465.9 & 12~~~
  &   ? &   ?~~~ &     ? &  ?~~~ \\                     
  & 2
  & 334 & 4.0~~~ & 553.9 & 15~~~
  &   ? &   ?~~~ &     ? &  ?~~~ \\                     
  & 3
  & 326 & 6.3~~~ & 691.8 & 18~~~
  &   ? &   ?~~~ &     ? &  ?~~~ \\                     
  & multi
  & 323 & 7.2~~~ & 1545.9 & 18~~~
  &   ? &   ?~~~ &    ? & ?~~~ \\                    
  \midrule
  &&&& \vdots \\
% \\
% level 1 &
% 387~ & 26.0~~~~~~ & 755.4~ &  18~~~
% \\
% &
% level 2 &
% 390~ & 25.4~~~~~~ & 1024.5~ &  20~~~
% \\
% &
% level 3 &
% 393~ & 24.9~~~~~~ & 7273.3~ &  25~~~
% \\
% &
% multi &
% 371~ & 29.1~~~~~~ & 9111.4~ &  21~~~
% \\
% \midrule
% \multirow{4}{*}{\makecell{\texttt{ltl2dgra}\\234 automata\\882 acc.~marks}}
% &
% level 1 &
% 552~ & 37.4~~~~~~ & 738.7~ & 14~~~
% \\
% &
% level 2 &
% 569~ & 35.5~~~~~~ & 907.3~ &  18~~~
% \\
% &
% level 3 &
% 561~ & 36.4~~~~~~ & 1163.4~ &  21~~~
% \\
% &
% multi &
% 542~ & 38.5~~~~~~ & 2320.7~ &  21~~~
% \\
% \midrule
% \multirow{4}{*}{\makecell{\texttt{ltl3tela}\\91 automata\\348 acc.~marks}}
% &
% level 1 &
% 332~ & 4.6~~~~~~ & 465.9~ &  12~~~
% \\
% &
% level 2 &
% 334~ & 4.0~~~~~~ & 553.9~ &  15~~~
% \\
% &
% level 3 &
% 326~ & 6.3~~~~~~ & 691.8~ &  18~~~
% \\
% &
% multi &
% 323~ & 7.2~~~~~~ & 1545.9~ & 18~~~
% \\
\bottomrule                                                                     
\end{tabular}
\end{table}

>>>>

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parametrized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 335 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G}) from the Spot library~\cite{duret.16.atva2}
and \texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA
by \texttt{ltl3tela} with option \texttt{-D1} and by another two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}. Some translators failed on some
formulas: it usually reached a timeout of 60 seconds or it produced an
automaton that cannot be parsed by the Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as
there is a little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions of the
translators. For each set of formulas and for each translator, it also
provides the number of fails, the number of produced automata with
less than two marks, and the number of automata with at least two
marks. The numbers of automata with at least two marks are typeset in
bold as these automata are actually used for the experimental
evaluation of our reduction technique.

@@

\begin{table}[t]
  \caption{The cumulative numbers of acceptance marks before and after
    reduction for various reduction levels and automata sets given by
    the translator and formula set. The column \emph{reduction} shows
    the cumulative percentage of saved acceptance marks and
    \emph{time} reports the cumulative reduction time. The column
    \textit{solver timeout} indicates the number of instances where
    the last call to the QBF solver timed out.  }
\label{tab:cumm-data}
\centering
\setlength{\tabcolsep}{3.9pt}
\newcommand{\mybox}[1]{\fbox{\begin{minipage}{4.7cm}\centering#1\end{minipage}}}
\begin{tabular}{lcrrrrrrrr}
  \toprule
  && \multicolumn{4}{c}{\makecell{reduction of marks in automata\\[-.8ex] from formulas from literature}}
  & \multicolumn{4}{c}{\makecell{reduction of marks in automata\\[-.8ex] from random formulas}}
  \\ \cmidrule(lr){3-6} \cmidrule(lr){7-10}
  translator & \makecell{reduction\\[-1ex] level}
  & marks & \makecell{reduction\\[-1ex] [\%]} & \makecell{time\\[-1ex] [s]} & \makecell{solver\\[-1ex] timeout} 
  & marks & \makecell{reduction\\[-1ex] [\%]} & \makecell{time\\[-1ex] [s]} & \makecell{solver\\[-1ex] timeout} \\
  \midrule
%  \multirow{5}{*}{\makecell{\texttt{ltl2tgba -G}}} &
  \multicolumn{2}{l}{\texttt{ltl2tgba -G}} 
  & \multicolumn{4}{c}{\mybox{198 marks in 70 automata}}
  & \multicolumn{4}{c}{\mybox{32 marks in 15 automata}}\\
  & 1
  & 198 & 0.0~~~ & 48.5 & 0~~~
  &   32 &   0.0~~~ &    8.4 & 0~~~ \\                     
  & 2
  & 198 & 0.0~~~ & 65.2 & 0~~~
  &   31 &   3.1~~~ &    9.4 & 0~~~ \\                     
  & 3
  & 189 & 4.5~~~ & 409.8 & 7~~~
  &   26 &   18.8~~~ &    43.9 & 1~~~ \\                     
  & multi
  & 189 & 4.5~~~ & 427.6 & 7~~~
  &   26 &   18.8~~~ &    44.9 & 1~~~ \\                    
  \midrule
%  \multirow{5}{*}{\makecell{\texttt{ltl2tgba -G}}} &
  \multicolumn{2}{l}{\texttt{ltl3tela}} 
  & \multicolumn{4}{c}{\mybox{348 marks in 91 automata}}
  & \multicolumn{4}{c}{\mybox{120 marks in 49 automata}}\\
  & 1
  & 332 & 4.6~~~ & 530.3 & 13~~~
  &   101 &   15.8~~~ &     32.4 &  0~~~ \\                     
  & 2
  & 334 & 4.0~~~ & 551.0 & 14~~~
  &   100 &   16.7~~~ &     32.3 &  0~~~ \\                     
  & 3
  & 326 & 6.3~~~ & 698.5 & 18~~~
  &   95 &   20.8~~~ &     66.9 &  1~~~ \\                     
  & multi
  & 319 & 8.3~~~ & 1619.2 & 18~~~
  &   95 &   20.8~~~ &    73.4 & 1~~~ \\                    
  \midrule
  %  \multirow{5}{*}{\makecell{\texttt{ltl2tgba -G}}} &
  \multicolumn{2}{l}{\texttt{ltl3tela -D1}} 
  & \multicolumn{4}{c}{\mybox{272 marks in 81 automata}}
  & \multicolumn{4}{c}{\mybox{97 marks in 44 automata}}\\
  & 1
  & 272 & 0.0~~~ & 383.1 & 9~~~
  &   95 &   2.1~~~ &     23.8 &  0~~~ \\                     
  & 2
  & 272 & 0.0~~~ & 386.6 & 10~~~
  &   95 &   2.1~~~ &     24.6 &  0~~~ \\                     
  & 3
  & 272 & 0.0~~~ & 950.2 & 14~~~
  &   92 &   5.2~~~ &     54.1 &  0~~~ \\                     
  & multi
  & 272 & 0.0~~~ & 1659.6 & 16~~~
  &   92 &   5.2~~~ &    67.2 & 1~~~ \\                    
  \midrule
  %  \multirow{5}{*}{\makecell{\texttt{ltl2tgba -G}}} &
  \multicolumn{2}{l}{\texttt{ltl2dgra}} 
  & \multicolumn{4}{c}{\mybox{882 marks in 234 automata}}
  & \multicolumn{4}{c}{\mybox{491 marks in 205 automata}}\\
  & 1
  & 544 & 38.3~~~ & 859.1 & 14~~~
  &   293 &   40.3~~~ &     275.6 &  0~~~ \\                     
  & 2
  & 554 & 37.2~~~ & 1073.5 & 17~~~
  &   280 &   43.0~~~ &     283.9 &  0~~~ \\                     
  & 3
  & 553 & 37.3~~~ & 1349.7 & 22~~~
  &   267 &   45.6~~~ &     433.6 &  3~~~ \\                     
  & multi
  & 535 & 39.3~~~ & 2434.0 & 23~~~
  &   264 &   46.2~~~ &    411.7 & 2~~~ \\                    
  \midrule
  %  \multirow{5}{*}{\makecell{\texttt{ltl2tgba -G}}} &
  \multicolumn{2}{l}{\texttt{ltl2dela}} 
  & \multicolumn{4}{c}{\mybox{348 marks in 91 automata}}
  & \multicolumn{4}{c}{\mybox{234 marks in 89 automata}}\\
  & 1
  & 332 & 4.6~~~ & 465.9 & 12~~~
  &   154 &   34.2~~~ &     89.0 &  0~~~ \\                     
  & 2
  & 334 & 4.0~~~ & 553.9 & 15~~~
  &   153 &   34.6~~~ &     123.6 &  0~~~ \\                     
  & 3
  & 326 & 6.3~~~ & 691.8 & 18~~~
  &   149 &   36.3~~~ &     172.7 &  2~~~ \\                     
  & multi
  & 323 & 7.2~~~ & 1545.9 & 18~~~
  &   148 &   36.8~~~ &    219.8 & 2~~~ \\

% \\
% level 1 &
% 387~ & 26.0~~~~~~ & 755.4~ &  18~~~
% \\
% &
% level 2 &
% 390~ & 25.4~~~~~~ & 1024.5~ &  20~~~
% \\
% &
% level 3 &
% 393~ & 24.9~~~~~~ & 7273.3~ &  25~~~
% \\
% &
% multi &
% 371~ & 29.1~~~~~~ & 9111.4~ &  21~~~
% \\
% \midrule
% \multirow{4}{*}{\makecell{\texttt{ltl2dgra}\\234 automata\\882 acc.~marks}}
% &
% level 1 &
% 552~ & 37.4~~~~~~ & 738.7~ & 14~~~
% \\
% &
% level 2 &
% 569~ & 35.5~~~~~~ & 907.3~ &  18~~~
% \\
% &
% level 3 &
% 561~ & 36.4~~~~~~ & 1163.4~ &  21~~~
% \\
% &
% multi &
% 542~ & 38.5~~~~~~ & 2320.7~ &  21~~~
% \\
% \midrule
% \multirow{4}{*}{\makecell{\texttt{ltl3tela}\\91 automata\\348 acc.~marks}}
% &
% level 1 &
% 332~ & 4.6~~~~~~ & 465.9~ &  12~~~
% \\
% &
% level 2 &
% 334~ & 4.0~~~~~~ & 553.9~ &  15~~~
% \\
% &
% level 3 &
% 326~ & 6.3~~~~~~ & 691.8~ &  18~~~
% \\
% &
% multi &
% 323~ & 7.2~~~~~~ & 1545.9~ & 18~~~
% \\
\bottomrule                                                                     
\end{tabular}
\end{table}
